---
title: "biodiv"
author: "Ilya"
date: "5/12/2018"
output: 
  github_document
always_allow_html: yes
---
Map species occurrence data from iNaturalist. iNaturalist (inaturalist.org) is a citizen science app. Assign iNaturalist data to U.S. parks. Make a map of number of species seen per park in NYC area. Data for NYC, color-coded by taxa, are available here:
https://fischhoff.shinyapps.io/shiny_test/ 
The data for the U.S. are here:
https://fischhoff.shinyapps.io/biodiversity/ (Note: loads slowly.)    

####install packages
```{r}
#install package dplyr if not already installed
list.of.packages <- c("dplyr", "leaflet", "shiny", "rgdal", "raster", "sp", "data.table", "htmltools")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
print(new.packages)
library(dplyr)
library(leaflet)
library(shiny)
library(rgdal)
library("raster") 
library("sp")
library(data.table)
library(htmltools)

```

####read in shapefile of protected areas and reproject and save as PAD_wgs84.Rdata
```{r}

#source: https://gapanalysis.usgs.gov/padus/data/download/ (national shapefile)
PAD <- shapefile("PADUS1_4Combined.shp")
#change datum
PAD_wgs84 <- spTransform(PAD, CRS("+proj=longlat +datum=WGS84"))

#assign cooordinates (package: sp)
PAD_coordinates = coordinates(PAD_wgs84)

PAD_wgs84$X = PAD_coordinates[,1]
PAD_wgs84$Y = PAD_coordinates[,2]

save(PAD_wgs84, file = "PAD_wgs84.Rdata")

```

####read in iNaturalist data 
```{r}
rm(list = ls())
library("raster")
library("data.table")
#occurrence download available here: 
#https://www.gbif.org/occurrence/download/0016851-180131172636756
inat = fread("occurrence.txt",blank.lines.skip=TRUE)
inat = as.data.frame(inat)
inat$longitude <- inat$decimalLongitude
inat$latitude <- inat$decimalLatitude
inat.ok= subset(inat, !is.na(decimalLatitude) & !is.na(decimalLongitude))
save(inat.ok, file = "inat.ok.Rdata")
```

####get park coordinates for inat data
```{r}
load("inat.ok.Rdata")
load("PAD_wgs84.Rdata")

#turn it into a SpatialPointsDataFrame
coordinates(inat.ok) <- c("decimalLongitude","decimalLatitude")#package sp

 # tell R that inat coordinates are in the same lat/lon reference system
# as the parks data 
projection(inat.ok) <- projection(PAD_wgs84)

# use 'over' with parks as a SpatialPolygonsDataFrame
# object, to determine which park (if any) contains each sighting, and
# store the park name as an attribute of the data
inat.ok$park <- over(inat.ok, PAD_wgs84)$Unit_Nm#name of park
inat.ok$X.park <- over(inat.ok, PAD_wgs84)$X
inat.ok$Y.park <- over(inat.ok, PAD_wgs84)$Y
inat.ok$park.acres <- over(inat.ok, PAD_wgs84)$GIS_Acres

save(inat.ok, file = "inat.ok.Rdata")

```

####summarize iNaturalist data by park: how many of each taxon are seen in each park?
```{r}
library("dplyr")
load("inat.ok.Rdata")
df = inat.ok
class_list = c("Amphibia", "Aves", "Mammalia", "Reptilia", "Arachnida", "Insecta", "Actinopterygii")
class_list_common = c("Amphibians", "Birds", "Mammals", "Reptiles", "Arachnids", "Insects", "Fish")
kingdom_list = c("Fungi", "Plantae")
organism_type <- function(df, class_list, class_list_common, kingdom_list) {
  df$organism_type = "Other"
  for (a in 1:length(class_list)){#for each class, assign that class to organism type
    df$organism_type[which(df$class==class_list[a])] = class_list_common[a]
  }
  for (b in 1:length(kingdom_list)){#for each kingdom, assign that kingdom to organism type
    df$organism_type[which(df$kingdom==kingdom_list[b])] = kingdom_list[b]
  }
  return(df)
}

ptm <- proc.time()#not long -- couple of seconds
df =organism_type(df, class_list, class_list_common, kingdom_list)
df = subset(df, organism_type!="Other")

keep.col = c("recordedBy",
             "X.park",
             "Y.park",
             "park",
             "park.acres",
             "species",
             "organism_type",
             "eventDate")
df =df[,keep.col]

ptm <- proc.time()#not long -- couple of seconds
df = data.frame(df)
df.park <- df %>%
  group_by(park, organism_type) %>%
    summarise(
    X.park = X.park[1],
    Y.park = Y.park[1],
    park.acres = park.acres[1],
    count = n(), 
    observers = length(unique(recordedBy)),
    species = length(unique(species))
    )
 # print(proc.time() - ptm)
#remove NA points
df.park = subset(df.park, !is.na(df.park$X.park))
save(df.park, file = "df.park.Rdata")
write.csv(df.park, file = "inat.park.csv")
```

#find the park with the most species (amphibians, all species, species by acre) in the U.S.
```{r}
load("df.park.Rdata")
A = subset(df.park,organism_type == "Amphibians")
Amax = subset(A, species == max(A$species))
print("park with most amphibians in U.S.")
print(Amax$park)
print(Amax$species)
#get number of total species by park
df.all <- df.park %>%
  group_by(park) %>%
  summarize(species_total = sum(species),
            species_per_acre = species_total/park.acres[1])
Max = subset(df.all, species_total == max(df.all$species_total))
print("park with most species")
print(Max$park)

Max.acre = subset(df.all, species_per_acre!="Inf")
Max.acre = subset(Max.acre, species_per_acre == max(Max.acre$species_per_acre))
print("park with most species per acre")
print(Max.acre$park)
print(Max.acre$species_per_acre)
```

####make a map for NYC area
```{r}
df= read.csv("inat.park.csv")
 df = filter(df,X.park >=-74.2589 & X.park< -73.7004 & Y.park <=40.9176 & Y.park>=40.4774)

df$X.park <- jitter(df$X.park, factor = 20)
df$Y.park <- jitter(df$Y.park, factor = 20) 
factpal <- colorFactor(topo.colors(9), df$organism_type)
# color.vec = c("#8c510a",#workaround for making each taxonomic group a different color
#                 "#bf812d",
#                 "#dfc27d",
#                 "#f6e8c3",
#                 "#f5f5f5",
#                 "#c7eae5",
#                 "#80cdc1",
#                 "#35978f",
#                 "#01665e")
  
    leaflet(df.park) %>%
      addTiles() %>%
       setView(lng = mean(df$X.park), lat = mean(df$Y.park) , zoom =10) %>%
      addCircles(lng = df$X.park,
                       lat = df$Y.park,
                 #color = color.vec[df$taxon_num],
                 #color = ~factpal(organism_type),
                        label = htmlEscape(paste(df$organism_type, df$species)))#label appears on hover
```


####run shiny app to make map online; requires separate files global.R, server.R
```{r}
library(shiny)
runApp()
#use rsconnect to connect to shinyapps.io (https://www.shinyapps.io/admin/#/tokens)
#use global.R, ui.R, server.R; 

```
